---
title: "Module 5 Review"
output: html_document
---

We will break into 4 groups. Each group will be assigned a problem to code and will write their answer(s) on the board. Refer to the two previoous lectures for help with code.

We will use the `GSE17260_eset` data set with genomic information for 110 advanced-stage serious ovarian cancer patients.  (A more thorough description can be found [here.](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0009615.))

Let's first upload the data using R:

```{r, message = FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("curatedOvarianData")
library("curatedOvarianData")
source(system.file("extdata", "patientselection.config", package = "curatedOvarianData"))
data("GSE17260_eset")
varLabels(GSE17260_eset)
set.seed(123)
```


### Problem 1

Predict the expression of the gene in the first column using the expressions in columns `2, 3,..., 11`. The data can be obtained using the command 

```{r}
mat.gene = exprs(GSE17260_eset) # Retrieve expression data from eSets.
```

1. Select the variables of interest
```{r}
# Your code here
```

2. Fit an exploratory linear regression model for the variable of interest `y`.

```{r}
# Your code here
```

3. Compute the $R^2$ summary.

```{r}
# Your code here
```


### Problem 2

This data set also provides the status of a patient, i.e. if the patient is alive or deceased, denoted by the variable `vital_status`. Predict the variable `vital_status` using the covariates `tumorstage` and `grade`, and then use these predictions to calculate the Brier score.

1. Select the variables from the data set:
```{r}
y = GSE17260_eset$vital_status
y = ifelse(y == "deceased", 1, 0)
```


2. Fit a logistic model to predict the variable `vital_status`.
```{r}
# Your code here
```


3. Compute the Brier Score.
```{r}
# Your code here
```



### Problem 3

Consider the same set of variables as in the previous exercise. Estimate the standard
deviation of the Brier Score using the bootstrap model.

```{r}
# Your code here
```



### Problem 4

Consider the variable `vital status` and use the same set of covariates as before combined
with the subset of genomic information `subset.gene` defined by

```{r}
subset.gene = t(mat.gene[1:50, ])
```

1. Fit a logistic model with this new set of predictors.

```{r}
# Your code here
```


2. Compute the AUC.

```{r}
# Your code here
```







